Survey of Cache Oblivious B-trees

### 2011-12-09 paper.tex
made more minor grammar edits

\title{Cache Oblivious B-trees}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\Insert{} works by initially inserting an element into the first-level array.
This array has size $2^0 = 1$, thus the element is in an array on it own.
Whenever you insert an element you check to see if there is another array of
the same size (this can be though of as checking to see if there is another
array in the same level). If there is another array with the same size, the
COLA merges the two together in sorted order. This merging will produce a new
array that is twice the size and thus goes into the next level. The COLA
continues to merge until there are no two arrays of the same size.

\Search{} can naively be performed by simply scanning each of the levels in
the COLA. Since everything is sorted the COLA could can use a binary search to
quickly find the element it is looking for. However, this is not very
efficient. As an improvement COLA, as the name suggest, uses look ahead
pointers to reduce the number of memory accesses required. Every 8th element
from the $(k+1)$st level exists in the $k$ level array in its proper sorted
location. This element also contains a pointer to its real location in the
$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
pointer''. Additionally, each every 4th cell in $k$th level array contains two
pointers pointing to the next and previous ``real lookahead pointers'' on this
level. The next and previous pointers are referred to as ``duplicate lookahead
pointers''. To search a level the COLA performs a binary search within an
appropriate range, determine the range in the next level using the ``duplicate
lookahead pointers'' and then recursively cascade up the tree.

@@

\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
based on the far that an element will never be merged more than $O(log(N))$
times. This is because at that point it will be in the last array.
Additionally, the first $log(B)$ merges that a element participates in uses a
constant number of memory transfer since we can assume that the relevant
blocks are already cached. Thus, we only need to worry about memory transfers
when we merge levels that have size greater than $B$. If the size of the array
is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
merging a single element is simply $1/B$. Since we just proved that an element
will only be merged $O(log(N))$ time, it follows that the overall amortized
memory transfers to insert an element is $O((log N)/B)$.

\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
show that the COLA never needs to search more than eight elements on each
level. Since there are $O(log(N))$ levels and since eight is a constant, it
follows that overall, search uses $O(log N)$ memory transfers. The fact that
only eight elements need to be search per level is shown through induction.
The base case is trivial. The first three levels of the COLA each have at most
eight elements (note that these elements are either actual real elements or
lookahead pointers). The COLA can obviously search these levels with no more
than eight member transfers. As the COLA follows lookahead pointers to the
higher levels, it will eventual get to a level where $k \geq 3$. Note that the
COLA does not ``enter'' this level at the start of the array; rather, its
arrived here by following the real lookahead pointers from the levels below
and thus is at some index $i$ that contains an element $e_i$ such that $e_i <
e$ and $e$ is the element being searched for. The goal at this point is to
find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
duplicate lookahead pointers can be used to locate the ``next'' and
``previous'' real lookahead pointers in the next level. The two real lookahead
pointers are used to and restrict the search in the next level. Recall that at
each level in the COLA every eighth element in a level is present in the
previous level. Thus, it follows that in next level only 8 elements need to be
searched. Furthermore, since the current level was reach from a lower level,
it follows that only 8 elements need to be search on this level to find the
appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
transfers.

@@

\section{Conclusion}

%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
Thanks Karger

>>>>

\title{Survey of Cache Oblivious B-trees}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\Insert{} works by initially inserting an element into the first-level array.
This array has size $2^0 = 1$, thus the element is in an array on it own.
Whenever you insert an element you check to see if there is another array of
the same size (this can be thought of as checking to see if there is another
array in the same level). If there is another array with the same size, the
COLA merges the two together in sorted order. This merging will produce a new
array that is twice the size and thus goes into the next level. The COLA
continues to merge until there are no two arrays of the same size.

\Search{} can naively be performed by simply scanning each of the levels in
the COLA. Since everything is sorted the COLA can use a binary search to
quickly find the element it is looking for. However, this is not very
efficient. As an improvement COLA, as the name suggests, uses look ahead
pointers to reduce the number of memory accesses required. Every 8th element
from the $(k+1)$st level exists in the $k$ level array in its proper sorted
location. This element also contains a pointer to its real location in the
$(k+1)$st level. Thus this pointer is referred to as ``real lookahead
pointer''. Additionally, every 4th cell in $k$th level array contains two
pointers pointing to the next and previous ``real lookahead pointers'' on this
level. The next and previous pointers are referred to as ``duplicate lookahead
pointers''. To search a level the COLA performs a binary search within an
appropriate range, determine the range in the next level using the ``duplicate
lookahead pointers'' and then recursively cascade up the tree.

@@

\Insert{} achieves an amortized $O(log N / B)$ memory transfers. The proof is
based on the fact that an element will never be merged more than $O(log(N))$
times. This is because at that point it will be in the last array.
Additionally, the first $log(B)$ merges that a element participates in uses a
constant number of memory transfers since we can assume that the relevant
blocks are already cached. Thus, we only need to worry about memory transfers
when we merge levels that have size greater than $B$. If the size of the array
is $k$ then the cost of merging is $O(k/B)$ and thus the amortized cost of a
merging a single element is simply $1/B$. Since we just proved that an element
will only be merged $O(log(N))$ time, it follows that the overall amortized
memory transfers to insert an element is $O((log N)/B)$.

\Search{} achieves $O(log N)$ memory transfers. The proof uses induction to
show that the COLA never needs to search more than eight elements on each
level. Since there are $O(log(N))$ levels and since eight is a constant, it
follows that overall, search uses $O(log N)$ memory transfers. The fact that
only eight elements need to be searched per level is shown through induction.
The base case is trivial. The first three levels of the COLA each have at most
eight elements (note that these elements are either actual real elements or
lookahead pointers). The COLA can obviously search these levels with no more
than eight member transfers. As the COLA follows lookahead pointers to the
higher levels, it will eventually get to a level where $k \geq 3$. Note that the
COLA does not ``enter'' this level at the start of the array; rather, its
arrived here by following the real lookahead pointers from the levels below
and thus is at some index $i$ that contains an element $e_i$ such that $e_i <
e$ and $e$ is the element being searched for. The goal at this point is to
find a new $i$ such that $e_i \leq e < e_{i+1}$. Once this $i$ is found the
duplicate lookahead pointers can be used to locate the ``next'' and
``previous'' real lookahead pointers in the next level. The two real lookahead
pointers are used to and restrict the search in the next level. Recall that at
each level in the COLA every eighth element in a level is present in the
previous level. Thus, it follows that in the next level only 8 elements need to be
searched. Furthermore, since the current level was reach from a lower level,
it follows that only 8 elements need to be searched on this level to find the
appropriate $i$. Thus, overall, the COLA uses $O(8log(N)) = O(log(N))$ memory
transfers.

@@

\section{Conclusion}
Fill this in.
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
Fill this in.
